RL-Musician: A Tool for Music Composition with Reinforcement Learning

### 2020-01-26 docs/paper/paper.tex
docs: Finalize draft of a paper

\begin{abstract}
In this paper, automatic composition of first species counterpoint is framed as an optimization problem close to reinforcement learning. An agent with shallow neural network as actor model composes polyphonic pieces sequentially one measure (bar) per time step. Once a specified number of measures is added, resulting piece is evaluated based on some hand-written rules derived from music theory. Cross-entropy method is used for searching weights of actor model that result in maximum expected score of a piece.

@@

Algorithmic music composition is automatic generation of outputs representing musical pieces and written in some formal notation. To name a few of common notations, there are sheet music, tablature, and MIDI standard. It is not required from output representation to unambiguously define sound waveform. For example, sheet music leaves exact loudness of played notes to discretion of a performer and may include only imprecise hints like pianissimo (very quiet). Anyway, there are parameters of sounds that must be determined by their representation. Usually, such parameters are pitch, start time, and duration.

@@

The word "species" in the name of the framework indicates that there are some types -- each of them has its own rhythmic pattern. Here, first species counterpoint is covered. This type has trivial rhythmic structure such that only whole (semibreve) notes can be used and every note must be placed in a beginning of a measure. Thus, interaction between rhythm and tonality is eliminated and the problem becomes even more simpler.

@@

Usually, softmax activation function is used in the last layer of generative RNN. However, it is a common practice in reinforcement learning domain to generalize it with Boltzmann softmax which is a family of activation functions parametrized by one parameter called temperature and denoted as $t$. Depending on temperature, output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$). In \cite{kumar2019polyphonic}, Boltzmann softmax is used and also input vector is extended by introducing an additional part indicating origination of initial input (so called plan). An environment is defined so that state is a pair of temperature and plan, actions are changes in either temperature or plan, and reward depends on evaluation of produced with these settings piece.

@@

In context of reinforcement learning, cross-entropy method is used for searching parameters of agent that result in maximum expected reward. So $X$ is a set of all possible finished episodes, $f$ is reward, and $u(\cdot, w)$ describes how an episode depends on parameters. Cross-entropy method is best suitable for environments where reward becomes known only after episode is finished. If there are rewards granted after intermediate steps, methods based on Q-values usually outperform it.

@@

Every line must have only small intervals between its successive elements in order to be perceived as a single line and not as a heterogeneous collection of sounds. Suppose that maximum allowed melodic interval is $s$ scale degrees (here, $s = 2$). If so, there are no more than $(2s + 1)^l$ options to fill the next measure given current measure. Actually, the number is smaller, because some options are filtered out by counterpoint rules. The rules enforced in this study can be either rules of voice leading (part writing) or rules of harmony.

Here, five rules of voice leading are imposed:
\begin{itemize}
	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated), because only they are stable;
	\item If a line has submediant followed by leading tone, tonic must be used after leading tone, because there is strong attraction to it; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad, because skip creates enough tension and something stable is needed;
	\item Movement in the opposite direction must happen immediately after a skip in order to resolve tension associated with the skip;
	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority), because resolution must be present during last measures.
\end{itemize}
In addition, there are two rules of harmony:
\begin{itemize}
	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant;
	\item For every pitch, the closest to it pitch from the same measure must be no more than tenth apart.
\end{itemize}

Above rules are absolute and an agent can not break them, because it selects pitches for the next measure only from options that are filtered according to them. If a set of such options is empty, episode terminates and an agent receives negative reward for dead end.

If the set of options is not empty, agent needs an observation. Observation is provided in terms of so called piano roll format. Denote size of pitch range available to the agent as $n$ and let the piece be represented also as a matrix $R$ of size $n \times m$ where $R_{ij}$ is 1 if $i$-th pitch is played in any line in $j$-th measure and 0 otherwise. Denote current measure as $t$, $1 \le t \le m$. Then observation is defined as vector $v$ of length $n$:
$$v = \sum_{i = 1}^t \beta^{t - i} R_{\cdot i},$$
where $\beta$ is a hyperparameter related to exponential decay, $0 < \beta < 1$.

For each option, agent concatenates observation $v$ with vector $u$ of length $(2s + 1)l$ encoding this option. Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $u$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them. Concatenation of $v$ and $u$ is passed as input to so called actor model which is a shallow neural network returning a single real-valued score. After these scores are collected for all valid options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions. An action is drawn from this distribution and the agents takes it.

Remember that episode starts with the first measure and the last measure already filled. The agent fills empty measures one-by-one according to described above procedure. When the final measure is reached, episode ends, created piece is evaluated and the agent receives reward.

@@

In this section, results are reported for experiments with 2 melodic lines and 16 measures starting from (G4, C5) and ending with (C4, C5). It is found that scores for autocorrelation and entropy of pitch distribution are redundant. Non-degenerate pieces are produced even without them. So only six properties are evaluated. There are weighted so that explicity of climax and absence of pitch class clashes have 3 times more impact than any other property.

In Table \ref{table:scores}, scores of a typical piece created by an agent trained for 20 populations, are compared with two benchmarks. The first one is simple baseline -- it is score of a piece created by an agent with weights that are the best amongst 100 random weights. The second one are scores for a piece manually created by the author.

\begin{table}[h!]
	\caption{Scores of sample pieces}
	\label{table:scores}
	\begin{center}
		\begin{tabular}{|l|c|c|c|c|}
			\hline
			& Baseline & Agent & Human & Maximum\\
			\hline
			Absence of looped pitches & -1 & 0 & 0 & 0\\
			\hline
			Absence of pitch or pitch class clashes & -0.429 & -0.429 & -0.643 & 0\\
			\hline
			Motion & -0.1 & 0.367 & 0.1 & 1\\
			\hline
			Correlation between lines & 0.552 & 0.599 & 0.278 & 1\\
			\hline
			Explicity of climax & 0.9 & 1.65 & 3 & 3\\
			\hline
			Number of skips & 0 & 1 & 1 & 1\\
			\hline
			Total reward & -0.076 & 3.187 & 3.735 & 6\\
			\hline
		\end{tabular}
	\end{center}
\end{table}

It can be seen that the agent's piece is better than the manually created piece by all metrics except explicity of climax. Since climax is higher than start note, both lines are assumed to have upward motion but the agent wrongly trades off climax score for contrary motion and lower lines correlation.

Table \ref{table:piece} contains the created by the agent piece for which above scores are granted.

\begin{table}[h!]
	\caption{A piece composed by an agent}
	\label{table:piece}
	\begin{center}
		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
			\hline
			& & & & & & & C6 & & & & & & & & \\
			\hline
			& & & & & & B5 & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & A5 & & & & A5 & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & G5 & & & & G5 & & G5 & G5 & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & F5 & & & & & & & & & & & & \\
			\hline
			& & E5 & & & & & & & & & & E5 & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& D5 & & & & & & & & & & & & & D5 & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			C5 & & & & & & & & & & & & & C5 & & C5 \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			G4 & & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& F4 & & & & & & & & & & & & & & \\
			\hline
			& & E4 & & & E4 & & E4 & & & E4 & & & E4 & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & D4 & & & D4 & & D4 & & & D4 & & & D4 & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & C4 & & & & & C4 & & & C4 & & & C4 \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

As for subjective evaluation of music quality, it is not correct to compare the selected piece created by the agent with the piece created by the author. The problem is that reward encourages to follow counterpointal rules but, from a modern listener point of view, breaking these rules does not necessarily sound unaesthetically. That being said, subjective perception of a resulting piece is not aligned with fitness function used during optimization. However, it must be noted that the piece shown above looks like overfitted to the given fitness function. It is not a problem associated with training, it is a flaw in problem setup.

@@

%feature representation.
%music theory: penalize downward skips and overlapping motion.

@@

The list of suggested further improvements is as follows:
\begin{itemize}
	\item \textbf{Reflect more knowledge from music theory in filtering rules and evaluational rules.} For example, overlapping motion should be discouraged. Also downward skips can be penalized in order to have upward skips before climax and downward stepwise motion after it.
	\item \textbf{Promote variety in fitness function.} Now, fitness function just averages rewards for several episodes. However, a new term can be added to this average. This term must reflect how different created pieces are from each other.
	\item \textbf{Revise feature representation for actor model.} Probably, procedure of next measure generation should be changed. Currently, it is assumed that shallow neural network is able to act based on relatively raw feature representation. To make it less raw, some features inspired by music theory may be engineered.
	\item \textbf{Abstract away all rules and focus only on essentials of music.} Tonal music introduces attraction between pitches. Depending on this attraction and degree of consonance/dissonance, tension increases or dissipates. Moments of time when level of tension changes may be rhythmically accented or unaccented. Close but not exact alignment between rhythmical accent and resolution results in aesthetic and interesting music. Taking into account voice leading constraints, optimization problem can be set up based on above considerations.
\end{itemize}

>>>>

\begin{abstract}
In this paper, automatic composition of first species counterpoint is framed as an optimization problem close to reinforcement learning. An agent with shallow neural network as an actor model composes polyphonic pieces sequentially one measure (bar) per time step. Once a specified number of measures is added, resulting piece is evaluated based on some hand-written rules derived from music theory. Cross-entropy method is used for searching weights of actor model that result in maximum expected score of a piece.

@@

Algorithmic music composition is automatic generation of outputs representing musical pieces and written in some formal notation. To name a few of common notations, there is sheet music, tablature, and MIDI standard. It is not required from output representation to unambiguously define sound waveform. For example, sheet music leaves exact loudness of played notes to discretion of a performer and may include only imprecise hints like pianissimo (very quiet). Anyway, there are parameters of sounds that must be determined by their representation. Usually, such parameters are pitch, start time, and duration.

@@

The word "species" in the name of the framework indicates that there are some types -- each of them has its own rhythmic pattern. Here, first species counterpoint is covered. This type has trivial rhythmic structure such that only whole (semibreve) notes can be used and every note must be placed in a beginning of a measure. Thus, interaction between rhythm and tonality is eliminated and the problem becomes even simpler.

@@

Usually, softmax activation function is used in the last layer of generative RNN. However, it is a common practice in reinforcement learning domain to generalize it with Boltzmann softmax which is a family of activation functions parametrized by one parameter called temperature and denoted as $t$. Depending on temperature, output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$). In \cite{kumar2019polyphonic}, Boltzmann softmax is used and also input vector is extended by introducing an additional part (so called plan) indicating origination of initial input. An environment is defined so that state is a pair of temperature and plan, actions are changes in either temperature or plan, and reward depends on evaluation of produced with these settings piece.

@@

In context of reinforcement learning, cross-entropy method is used for searching parameters of agent that result in maximum expected reward. So $X$ is a set of all possible finished episodes, $f$ is reward, and $u(\cdot, w)$ describes how an episode depends on parameters. Cross-entropy method is best suitable for environments where reward becomes known only after an episode is finished. If there are rewards granted after intermediate steps, methods based on Q-values usually outperform it.

@@

Every line must have only small intervals between its successive elements in order to be perceived as a single line and not as a heterogeneous collection of sounds. Suppose that maximum allowed melodic interval is $s$ scale degrees (here, $s = 3$). If so, there are no more than $(2s + 1)^l$ options to fill the next measure given current measure. Actually, the number is smaller, because some options are filtered out by counterpoint rules. The rules enforced in this study can be either rules of voice leading (part writing) or rules of harmony.

Here, five rules of voice leading are imposed:
\begin{itemize}
	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated), because only they are stable;
	\item If a line has submediant followed by leading tone, tonic must be used after leading tone, because there is strong attraction to it; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
	\item If a melodic interval between two successive pitches from the same line is larger than second (i.e., if there is a skip), the second pitch must be from tonic triad, because skip creates enough tension and something stable is needed;
	\item Movement in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
	\item For every pitch, there must be a way to reach the final pitch of this line with step motion (the final measure is filled initially with user-defined sonority), because resolution must be present during last measures.
\end{itemize}
In addition, there are two rules of harmony:
\begin{itemize}
	\item All harmonic intervals between two simultaneously sounding pitches from different lines must be consonant;
	\item For every pitch, the closest to it pitch from the same measure must be no more than tenth apart.
\end{itemize}

Above rules are absolute and an agent cannot break them, because it selects pitches for the next measure only from options that are filtered according to them. If a set of such options is empty, episode terminates and an agent receives negative reward for dead end.

If the set of options is not empty, an agent needs an observation. Observation is provided in terms of so called piano roll format. Denote size of pitch range available to the agent as $n$ and let the piece be represented also as a matrix $R$ of size $n \times m$ where $R_{ij}$ is 1 if $i$-th pitch is played in any line in $j$-th measure and 0 otherwise. Denote current measure as $t$, $1 \le t \le m$. Then observation is defined as vector $v$ of length $n$:
$$v = \sum_{i = 1}^t \beta^{t - i} R_{\cdot i},$$
where $\beta$ is a hyperparameter related to exponential decay, $0 < \beta < 1$.

For each option, agent concatenates observation $v$ with encoding this option vector $u$ of length $(2s + 1)l$. Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $u$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them. Concatenation of $v$ and $u$ is passed as input to so called actor model which is a shallow neural network returning a single real-valued score. After these scores are collected for all valid options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions. An action is drawn from this distribution and the agent takes it.

Remember that an episode starts with the first and last measures already filled. The agent fills empty measures one-by-one according to the described above procedure. When the final measure is reached, the episode ends, created piece is evaluated and the agent receives reward.

@@

In this section, results are reported for experiments with 2 melodic lines and 16 measures starting from (G4, C5) and ending with (C4, C5). It is found that scores for autocorrelation and entropy of pitch distribution are redundant. Non-degenerate pieces are produced even without them. So only six properties are evaluated. They are weighted so that explicity of climax and absence of pitch class clashes have 3 times more impact than any other property.

In Table \ref{table:scores}, scores of a typical piece created by an agent trained for 20 populations, are compared with two benchmarks. The first one is simple baseline -- it is score of a piece created by an agent with weights that are the best amongst 100 random weights. The second one are scores for a piece manually created by the author in about 15 minutes.

\begin{table}[h!]
	\caption{Scores of sample pieces}
	\label{table:scores}
	\begin{center}
		\begin{tabular}{|l|c|c|c|c|}
			\hline
			& Baseline & Agent & Human & Maximum\\
			\hline
			Absence of looped pitches & -1 & 0 & 0 & 0\\
			\hline
			Absence of pitch or pitch class clashes & -0.429 & -1.286 & -0.643 & 0\\
			\hline
			Motion & -0.1 & -0.033 & 0.1 & 1\\
			\hline
			Correlation between lines & 0.552 & 0.168 & 0.278 & 1\\
			\hline
			Explicity of climax & 0.9 & 3 & 3 & 3\\
			\hline
			Number of skips & 0 & 1 & 1 & 1\\
			\hline
			Total reward & -0.076 & 2.849 & 3.735 & 6\\
			\hline
		\end{tabular}
	\end{center}
\end{table}

It can be seen that the agent's piece consists of two lines with formally good melodies but these lines are not independent enough. Table \ref{table:piece} contains the created by the agent piece for which above scores are reported. Visual analysis of the shown piano roll confirms observation about lines dependence -- their climax points coincide and there are plenty of intermediate octave intervals.

\begin{table}[h!]
	\caption{A piece composed by an agent}
	\label{table:piece}
	\begin{center}
		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
			\hline
			& & & & & & & C6 & & & & & & & & \\
			\hline
			& & & & & & B5 & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & A5 & & & & A5 & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & G5 & & & & G5 & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & F5 & & & & & & & & F5 & & & & \\
			\hline
			& & E5 & & & & & & & & E5 & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& D5 & & & & & & & & & & & & D5 & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			C5 & & & & & & & C5 & & & & & C5 & & C5 & C5 \\
			\hline
			& & & & & & & & B4 & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & A4 & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			G4 & & & & G4 & & G4 & & & & G4 & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& F4 & & & & F4 & & & & & & F4 & & & & \\
			\hline
			& & & & & & & & & & & & E4 & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & D4 & & & & & & & & & & D4 & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & C4 & & & & & & & & & & & & C4 & C4 \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

As for subjective evaluation of music quality, it is not correct to compare the selected piece created by the agent with the piece created by the author. The problem is that reward encourages to follow counterpointal rules but, from a modern listener point of view, breaking these rules does not necessarily sound unaesthetically. That being said, subjective perception of a resulting piece is not aligned with fitness function used during optimization. However, it must be noted that the piece shown above looks like excessively oriented on maximization of fitness function regardless more subtle properties important for human listeners. It is not a problem associated with training, it is a flaw in problem setup (more criteria for evaluation are needed).

@@

@@

The list of suggested further improvements is as follows:
\begin{itemize}
	\item \textbf{Reflect more knowledge from music theory in filtering rules and evaluational rules.} For example, overlapping motion should be discouraged. Also downward skips can be penalized in order to have upward skips before climax and downward conjunct motion after it.
	\item \textbf{Promote variety in fitness function.} Now, fitness function just averages rewards for several episodes. However, a new term can be added to this average. This term must reflect how different created pieces are from each other.
	\item \textbf{Revise feature representation for actor model.} Probably, procedure of next measure generation should be changed. Currently, it is assumed that shallow neural network is able to act based on relatively raw feature representation. To make it less raw, some features inspired by music theory may be engineered.
	\item \textbf{Tune hyperparameters.} In the current software implementation, cross-entropy method has five hyperparameters, an agent has two hyperparameters, and evaluation functions have more than ten hyperparameters. As of now, not so many computational experiments are held in order to find their optimum values. 
	\item \textbf{Abstract away all rules and focus only on essentials of music.} Tonal music introduces attraction between pitches. Depending on this attraction and degree of consonance/dissonance, tension increases or dissipates. Moments of time when level of tension changes may be rhythmically accented or unaccented. Close but not exact alignment between rhythmical accent and resolution often results in aesthetic and interesting music. Taking into account constraints of smooth voice leading, optimization problem can be set up based on above considerations.
\end{itemize}
